### 2015-03-18 doc/thesis/evaluation.tex
Minor improvements to the evaluation results

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
Lua Standard Library & 62\% & 6\% & 9\% & 23\% & 129 \\ % 100%
\hline
MD5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
\hline
LuaSocket & 89\% & 1\% & 2\% & 8\% & 123 \\ % 100%
\hline
LuaFileSystem & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
\hline
HTTP Digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
\hline
Typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
Modulo 11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
\hline
Typed Lua Compiler & 93\% & 0\% & 1\% & 6\% & 154 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

Table \ref{tab:evalbycase} summarizes our evaluation results for each
of the case studies that we used Typed Lua for typing their members.
We split the members of each case study into four categories:
\emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
The \emph{easy} category shows the percentage of members that
we could give a precise static type.
The \emph{poly} category shows the percentage of members that
we made minimal use of the dynamic type as a replacement for the
lack of type parameters.
The \emph{over} category shows the percentage of members that
we gave a static type that is not precise enough,
as these members are overloaded functions that require intersection
types to describe their precise static types.
The \emph{hard} category shows the percentage of members that
we had to rely on the dynamic type because they rely on reflection.
The last column of the table shows the total number of members
of each case study.
The results that we obtained for the \emph{easy} category give
a lower bound on how much static type safety
we could add to each one of our case studies.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
base & 35\% & 12\% & 8\% & 45\% & 26 \\ % 100%
\hline
coroutine & 14\% & 0\% & 0\% & 86\% & 7 \\ % 100%
\hline
package & 62\% & 0\% & 0\% & 38\% & 8 \\ % 100%
\hline
string & 75\% & 0\% & 0\% & 25\% & 16 \\ % 100%
\hline
utf8 & 100\% & 0\% & 0\% & 0\% & 6 \\ % 100%
\hline
table & 14\% & 72\% & 14\% & 0\% & 7 \\ % 100%
\hline
math & 81\% & 0\% & 19\% & 0\% & 27 \\ % 100%
\hline
io & 71\% & 0\% & 10\% & 19\% & 21 \\ % 100%
\hline
os & 82\% & 0\% & 18\% & 0\% & 11 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Library}
\label{tab:evallsl}
\end{table}

@@

There are some functions in the \texttt{base} module that
we could not give a precise static type because our type
system does not have parametric polymorphism,
as the \emph{poly} category shows.
This is the case of \texttt{ipairs}, \texttt{rawget}, and
\texttt{rawset}.

@@

In the case of \texttt{tonumber}, the type of the first parameter
depends on the type of the second parameter.
For instance, we can call \texttt{tonumber(1)}, but we cannot
call \texttt{tonumber(1,2)}.
More precisely, the first argument of \texttt{tonumber} can be
a value of any type if it is the only argument, but it must
be a string if there is a second argument,
which must be an integer.

@@

The \texttt{coroutine} module was also very difficult to type
because our type system cannot describe the computational effects
of a program.
The \emph{hard} category shows the amount of functions that we
could not give a precise static type for this reason.
Lua has one-shot delimited continuations \citep{james2011yield}
in the form of \emph{coroutines} \citep{moura2009rc}, and
effect systems \citep{nielson1999type} are an approach that we
could use to describe control transfers with continuations.
However, for now, coroutines are out of the scope of our type
system, and we use an empty \emph{userdata} declaration
to represent the type \emph{thread}.

@@

We could give precise static types to the constants and functions
of the \texttt{package} module, but we could not give precise static
to the following tables:
\texttt{package.loaded}, \texttt{package.preload}, and \texttt{package.loaders}.
The first stores loaded modules, while the others store module \emph{loaders}.
They are difficult to type because their types rely on reflection,
that is, their types depend on the modules a program loads.
For this reason, they are in the \emph{hard} category.

@@

Lua programmers often use the functions of the \texttt{string} module
in an object-oriented style.
For instance, \texttt{string.len("foo")} and \texttt{("foo"):len()}
produce the same result.
We did not include the typing of the \texttt{string} methods in our
evaluation results, because we treat this feature as a special case
in the implementation.

@@

Even if our type system had parametric polymorphism, there is
still one function of the \texttt{table} module that we could
not give a precise static type because it is an overloaded function,
as the \emph{over} category shows.
This function is \texttt{table.insert}, and its type depends on
the calling arity.
More precisely, calling \texttt{table.insert(l, v)} inserts the
element \texttt{v} at the end of the list \texttt{l},
while \texttt{table.insert(l, p, v)} inserts the element \texttt{v}
at the position \texttt{p} of the list \texttt{l},
but generates a run-time error when \texttt{p} is out of bounds.
This function also does not follow the semantics of Lua on
discarding extra arguments, and generates a run-time error whenever
we pass more than three arguments, even if the first three arguments
match its signature.

@@

In the case of \texttt{math.abs},
calling \texttt{math.abs(1)} returns the integer \texttt{1},
while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.
Even though we gave this function the type \texttt{(number) -> (number)},
it is not precise enough because the return type is always \texttt{number}
independently of the argument type.
In other words, the return type should be \texttt{integer} when
the argument type is also \texttt{integer}, and the return type
should be \texttt{number} when the argument type is also \texttt{number}.

In the case of \texttt{math.random}, its type depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
Like \texttt{table.insert}, this function also does not follow the
semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.

The \texttt{io} module provides operations for manipulating files,
and these operations can use implicit or explicit file descriptors.
The implicit operations are functions in the \texttt{io} table,
while the explicit operations are methods of a file descriptor.
We used an \emph{userdata} declaration to introduce the type
\texttt{file} for representing the type of a file descriptor
and its methods.
The evaluation results include both implicit and explicit operations.

We could give precise static types to most of the members of the
\texttt{io} module, but its categories \emph{over} and \emph{hard}
show that it includes several members that we could not give a precise
static type.
While the function \texttt{io.close} and the method \texttt{file:close}
are in the \emph{over} category,
the functions \texttt{io.read} and \texttt{io.lines} are
in the \emph{hard} category along with the methods
\texttt{file:read} and \texttt{file:lines}.

We could not precisely type \texttt{io.close} because it is
an overloaded function.
This means that its return type depends on whether the file handle
that is being closed was created with \texttt{io.popen}.
Calling \texttt{io.close(f)} returns the tuple \texttt{(boolean?, string, number)}
when \texttt{f} was created with \texttt{io.popen},
but it returns the tuple \texttt{(boolean)?} otherwise.
The method \texttt{file:close} has the same issue.

@@

The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
The \texttt{mime} module offers low-level and high-level filters
that apply and remove some text encoding.
The low-level filters are written in C, while the high-level filters
use the function \texttt{ltn12.filter.cycle} along with the low-level
filters to create standard filters.

@@

LuaFileSystem is a library that extends the set of functions
for manipulating file systems in Lua.
It contains just the \texttt{lfs} module that is written in C,
so we used Typed Lua's description file to type it.
Table \ref{tab:evallfs} summarizes the evaluation results for LuaFileSystem.

@@

Even though we could not precisely type the interface that \texttt{http-digest}
exports, we could use only static types to annotate it, and they pointed a bug
in the code.
The problem was related to the way the library was loading the MD5 library
that should be used. 
This part of the code checks the existence of three different MD5 libraries
in the system, and uses the first one that is available, or generates an
error when none is available.
The code that loads the first option was fine, but the code that loads the
second and third options were trying to access an undefined global variable.

@@

Third, the module uses \texttt{ipairs} to iterate over an array of functions,
but our type system also has limited support to \texttt{ipairs}, and generates
a warning when we try to use the indexed value inside the \texttt{for} body.
As we mentioned in this chapter, we use the dynamic type as a replacement for
the lack of type parameters.
This means that we get warnings inside an \texttt{ipairs} iteration because
all iterated elements have the dynamic type.
We removed this warning using the numeric \texttt{for} to perform the same loop.

@@

The interface that the \texttt{tlchecker} and \texttt{tlcode}
modules export are also straightforward to type.
The former traverses the AST to perform type checking,
while the latter traverses the AST to perform code generation.

Even though we could precisely type the interface that most of
the modules export, we had issues to write mutually recursive functions.
This kind of functions often appear in compilers construction
to traverse the data structures that they use.
However, Typed Lua still does not support mutually recursive functions.
A way to overcome this limitation was to predeclare these functions
with an empty body, and then redeclare them with they actual body.
The first declaration specifies the function type, while the
second specifies what the function actually does without changing
any type definition.

Traversing the AST would also be problematic if we had not included
a way to discriminate unions of table types, as we mentioned in Section \ref{sec:alias}.

>>>>

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
Lua Standard Library & 64\% & 5\% & 8\% & 23\% & 129 \\ % 100%
\hline
MD5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
\hline
LuaSocket & 89\% & 1\% & 2\% & 8\% & 123 \\ % 100%
\hline
LuaFileSystem & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
\hline
HTTP Digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
\hline
Typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
Modulo 11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
\hline
Typed Lua Compiler & 93\% & 0\% & 1\% & 6\% & 154 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

Table \ref{tab:evalbycase} summarizes our evaluation results for each
of the case studies that we used Typed Lua for typing their members.
An exported member is any Lua value that a module might export.
We split the members of each case study into four categories:
\emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
The \emph{easy} category shows the percentage of members that
we could give a precise static type.
The \emph{poly} category shows the percentage of members that
we made minimal use of the dynamic type as a replacement for the
lack of type parameters.
The \emph{over} category shows the percentage of members that
we gave a static type that is not precise enough,
as these members are overloaded functions that require intersection
types to describe their precise static types.
The \emph{hard} category shows the percentage of members that
we could not give a precise static type.
The last column of the table shows the total number of members
of each case study.
The results that we obtained for the \emph{easy} category give
a lower bound on how much static type safety
we could add to each one of our case studies.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
base & 35\% & 4\% & 8\% & 53\% & 26 \\ % 100%
\hline
coroutine & 14\% & 0\% & 0\% & 86\% & 7 \\ % 100%
\hline
package & 62\% & 0\% & 0\% & 38\% & 8 \\ % 100%
\hline
string & 75\% & 0\% & 0\% & 25\% & 16 \\ % 100%
\hline
utf8 & 100\% & 0\% & 0\% & 0\% & 6 \\ % 100%
\hline
table & 14\% & 72\% & 14\% & 0\% & 7 \\ % 100%
\hline
math & 81\% & 0\% & 19\% & 0\% & 27 \\ % 100%
\hline
io & 81\% & 0\% & 0\% & 19\% & 21 \\ % 100%
\hline
os & 82\% & 0\% & 18\% & 0\% & 11 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Library}
\label{tab:evallsl}
\end{table}

@@

There are some functions in the \texttt{base} module that
we could not give a precise static type because our type
system does not have parametric polymorphism,
as the \emph{poly} category shows.
This is the case of \texttt{ipairs}.

@@

In the case of \texttt{tonumber}, it has two different types:
\texttt{(value) -> (number)} and \texttt{(string, integer) -> (number)}.
This means that the type of the first parameter depends on the
type of the second parameter.
For instance, we can call \texttt{tonumber(1)}, but we cannot
call \texttt{tonumber(1,2)}.
Note that the first argument of \texttt{tonumber} can be
a value of any type if it is the only argument, but it must
be a string if there is a second argument,
which must be an integer.

@@

The \texttt{coroutine} module was also very difficult to type
because our type system cannot describe the computational effects
of a program.
The \emph{hard} category shows the amount of functions that we
could not give a precise static type for this reason.
Lua has one-shot delimited continuations \citep{james2011yield}
in the form of \emph{coroutines} \citep{moura2009rc}, and
effect systems \citep{nielson1999type} are an approach that we
could use to describe control transfers with continuations.
However, for now, coroutines are out of scope of our type
system, and we use an empty \texttt{userdata} declaration
to represent the type \texttt{thread}.

@@

We could give precise static types to the constants and functions
of the \texttt{package} module, but we could not give precise static
types to the following tables that it exports:
\texttt{package.loaded}, \texttt{package.preload}, and \texttt{package.loaders}.
The first stores loaded modules, while the others store module \emph{loaders}.
They are difficult to type because their types rely on reflection,
that is, their types depend on the modules a program loads.
For this reason, they are in the \emph{hard} category.

@@

@@

Even if our type system had parametric polymorphism, there is
still one function of the \texttt{table} module that we could
not give a precise static type because it is an overloaded function,
as the \emph{over} category shows.
This function is \texttt{table.insert}, and its type depends on
the calling arity.
That is, calling \texttt{table.insert(l, v)} inserts the
element \texttt{v} at the end of the list \texttt{l},
while \texttt{table.insert(l, p, v)} inserts the element \texttt{v}
at the position \texttt{p} of the list \texttt{l},
and generates a run-time error when \texttt{p} is out of bounds.
This function also does not follow the semantics of Lua on
discarding extra arguments, and generates a run-time error whenever
we pass more than three arguments, even if the first three arguments
match its signature.

@@

In the case of \texttt{math.abs},
it has two different types: \texttt{(integer) -> (integer)} and
\texttt{(number) -> (number)}.
Even though we gave this function the more general type \texttt{(number) -> (number)},
it is not precise enough because the return type is always \texttt{number}
independently of the argument type.
In other words, the return type should be \texttt{integer} when
the argument type is also \texttt{integer}, and the return type
should be \texttt{number} when the argument type is also \texttt{number}.

In the case of \texttt{math.random}, it also has two different types:
\texttt{() -> (number)} and \texttt{(integer, integer?) -> (integer)}.
This means that the type of \texttt{math.random} depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
Like \texttt{table.insert}, this function also does not follow the
semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.

The \texttt{io} module provides operations for manipulating files,
and these operations can use implicit or explicit file descriptors.
The implicit operations are functions in the \texttt{io} table,
while the explicit operations are methods of a file descriptor.
We used an \texttt{userdata} declaration to introduce the type
\texttt{file} for representing the type of a file descriptor
and its methods.
The evaluation results include both implicit and explicit operations.

We could give precise static types to most of the members of the
\texttt{io} module, but the \emph{hard} category shows that it
includes some members that we could not give a precise static type.
The functions \texttt{io.read} and \texttt{io.lines} are
in the \emph{hard} category along with the methods
\texttt{file:read} and \texttt{file:lines}.

@@

The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
The \texttt{mime} module offers low-level and high-level filters
that apply and remove some text encodings.
The low-level filters are written in C, while the high-level filters
use the function \texttt{ltn12.filter.cycle} along with the low-level
filters to create standard filters.

@@

LuaFileSystem is a library that extends the set of functions
for manipulating file systems in Lua.
It contains just the \texttt{lfs} module that is written in C,
so we used Typed Lua's description files to type it.
Table \ref{tab:evallfs} summarizes the evaluation results for LuaFileSystem.

@@

Even though we could not precisely type the interface that \texttt{http-digest}
exports, we could use only static types to annotate this module,
and they pointed a bug in the code.
The problem was related to the way the library was loading the MD5 library
that should be used. 
This part of the code checks the existence of three different MD5 libraries
in the system, and uses the first one that is available, or generates an
error when none is available.
The code that loads the first option was fine, but the code that loads the
second and third options were trying to access an undefined global variable.

@@

Third, the module uses \texttt{ipairs} to iterate over an array of functions,
but our type system also has limited support to \texttt{ipairs}, and generates
a warning when we try to use the indexed value inside the \texttt{for} body.
As we mentioned in this chapter, we use the dynamic type as a replacement for
the lack of type parameters.
This means that we get warnings inside an \texttt{ipairs} iteration, because
all iterated elements have the dynamic type.
We removed this warning using the numeric \texttt{for} to perform the same loop.

@@

We could type the interfaces that modules \texttt{tlchecker} and
\texttt{tlcode} export.
The former traverses the AST to perform type checking,
while the latter traverses the AST to perform code generation.

Even though we could precisely type the interface that most of
the modules export, we had issues to write mutually recursive functions.
This kind of functions often appear in compilers construction
to traverse the data structures that they use.
However, Typed Lua still does not support mutually recursive functions.
A way to overcome this limitation was to predeclare these functions
with an empty body, and then redeclare them with their actual body.
The first declaration specifies the function type, while the
second specifies what the function actually does without changing
any type definition.

Traversing the AST would also be problematic if we had not included
a way to discriminate unions of table types, as we mentioned in Section \ref{sec:alias}.
Without a way to discriminate a union of table types, any attempt
to index this union of table types would generate a warning.
